---
title: "introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(r4msigdb)
```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# r4msigdb

[MSigDB](https://www.gsea-msigdb.org/gsea/msigdb/) is a widely used gene set database in bio-research. However, navigating and querying pathways of interest on its website can be challenging. This R package is designed to facilitate more convenient and efficient querying of pathways either based on specific genes or using regular expression patterns to match pathway names.

# Installation

You can install the development version of r4msigdb like so:

```{r, eval=FALSE}
devtools::install('snowGlint/r4msigdb')
```

# Querying Pathways

## Search for Pathways Related to a Specific Topic

To retrieve pathways related to a specific topic (e.g., cell programmed death):

```{r example1}
library(r4msigdb)
query(species = 'Hs', pathway = 'OPTOSIS') |> head()
```

This will return a list of pathways associated with `OPTOSIS` and the genes involved in each pathway.

## Search for Pathways Related to Specific Genes

If you want to find pathways related to specific genes (e.g., PTPRC and TP53):

```{r example2}
query(species = 'Hs', symbols = c('PTPRC', 'TP53')) |> head()
```

This will provide pathways where the specified genes are involved.

## Custom Query

Advanced users familiar with data.table can perform custom queries. For example, to retrieve all pathways `collection_name == 'H'`:

```{r}
query(species = 'Hs')[.('H')] |> head()
```

You can also use `.unlist = TRUE` to unlist the symbols column in any of the above query methods.

```{r}
query(species = 'Hs', .unlist = TRUE) |> head()
```

# GSEA

```{r, eval=FALSE}
library(fgsea)
library(data.table)
palette <- c("#440154FF", "#31688EFF", "#26828EFF", "#6DCD59FF", "#FDE725FF")

data(exampleRanks)
pathway <-  query(species = 'Hs', pathway = 'OPTOSIS', .unlist = TRUE)
pathway <- pathway[, .(standard_name, symbol)]
# substitute names with random symbols
set.seed(2024)
names(exampleRanks) <- query(species = 'Hs', .unlist = TRUE)[, sample(unique(symbol), length(exampleRanks))]

gseaR <- clusterProfiler::GSEA(rev(exampleRanks),
  TERM2GENE = pathway, pvalueCutoff = 1, by = "fgsea", eps = 0
)
sortedgsea <- as.data.table(gseaR@result)[order(pvalue)]

enrichplot::gseaplot2(gseaR, sortedgsea[["ID"]][1:5],
  base_size = 10,
  color = palette,
  rel_heights = c(1.5, 0.3, 0.5),
  pvalue_table = FALSE
)
```
